Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2014-10-06 draft_lm1.tex
added more figures
fixed bad link for PNG fig
worked on the introduction

\section*{Introduction}
When modeling infecctious disease spread using dynamic models, the initial 
conditions of the system, although key to match the simulated dynamics to 
observed data, are frequently overlooked.
If the model is about a disease 
invading a new population one can simply assume the vast majority of the 
population to be susceptible.
However for a disease which is well in its way to 
endemicity but not yet stable, it can be very hard to determine the 
immunological structure of the population without resorting to expensive large 
scale prevalence tests.

In the particular case of Dengue fever in Brazil, with four circulating 
serotypes, it can be very challenging to parameterize a simulation starting at 
any time other than when the date the disease was reintroduced in the country 
in the late 1980s.
For SIR-like models, the relevant initial values are the 
number of infectious and susceptible  individuals. 
The number of infectious 
individuals are less challenging to determine, because the get very small 
during the winters and can be derived from incidence data  with is monitored 
continuously.
The real challenge is to determine the number of susceptibles to 
a particular strain right at the beginning of the dengue season.

Multiple factors contribute to the dificulty in estimating the number of 
susceptibles in the population, first and foremost is the fact that dengue

confers permanente immunity within serotypes but with only short-lived 
cross-immunity. Secondly, there are a larger number of asymptomatic cases in 
any epidemic, which nevertheless acquire immunity, Thirdly, under-reporting is 
a serious issue in Brazil. Duarte and Fran\c{c}a\cite{duarte_data_2006}, estimated 
the sensitivity of dengue reporting in Belo-Horizonte, Brazil to be of 63\%, 
meaning that aproximately 37\% of the cases go unreported. Lastly, demography 
and migrations affect the number of susceptible in ways which are not easy to
determine.

The initial number of susceptibles -- $S_0$ -- is a a very powerful parameter 
with regard to epidemic risk.
Its value affects directly the local $R_0$, 
thus being intrumental in predicting epidemics, but it also affects the speed 
and size of the epidemic.

In this paper, we estimate the number of susceptibles, allowing for 
discontinuities right before every epidemic in the last 18 years. THis way we 
can account for replacement of the predominant serotype, which can redefine 
susceptibility for the following season.

@@

\bibliography{lm1}
\section*{Figure Legends}
% This section is for figure legends only, do not include
% graphics in your manuscript file.
%
%\begin{figure}
%\caption{
%{\bf Bold the first sentence.}  Rest of figure caption.  
%}
%\label{Figure_label}
%\end{figure}

@@

>>>>

\section*{Introduction}
When modeling infecctious disease spread using dynamic models, the initial 
conditions of the system, although key to match the simulated dynamics to 
observed data, are frequently overlooked.

Dengue is a disease which is caused by 4 different types of viruses 
which co-circulate between the vector and human populations. immunity is 
considered permanent after the exposure of humans to each of the viruses but 
cross immunity is considered very limited[ref].
Thus, the proportions of viral types co-circulating at any point in time is 
strongly dependent on previous incidence of the disease, which determines the 
number of susceptibles for each viral type at any point in time.

Dengue transmission is conditioned on temperature, due to its effects on the 
vector reproduction.
In places with sufficient temperature variations dengue is a summer disease 
mainly.

Determining the viral types in circulation during an epidemic is 
expensive, as it requires virus isolatio and reverse-transcriptase-PCR. 
Therefore only a very limited number of samples are examined on any epidemic. 
Besides, mild cases are rarely detected let alone tested for dengue.

Dengue in Brazil is a disease which is well in its way to 
endemicity but not yet stable, therefore co-circulation patterns have not yet 
reached a steady state as has been demonstrated for certain countries in 
south-east asia which have had the disease much longer.

In the particular case of Dengue fever in Brazil, with four circulating 
serotypes, it can be very challenging to parameterize a simulation starting at 
any time other than when the date the disease was reintroduced in the country 
in the late 1980s and even so, without virological survey data from time to 
time, to inform on the mixing of viruses in circulation, it is virtually 
impossible to predict, by simulation alone, the imunological state of the 
population at any point in the future.

For SIR-like models, the relevant initial values are the number of infectious 
and susceptible  individuals. 
The number of infectious individuals are less challenging to determine, because 
the get very small during the winters and can be derived from incidence data  
with is monitored continuously.
The real challenge is to determine the number of susceptibles to 
a particular strain right at the beginning of the dengue season.

Multiple factors contribute to the dificulty in estimating the number of 
susceptibles in the population, first and foremost is the fact that dengue
confers permanent immunity within serotypes but with only short-lived 
cross-immunity. 
Secondly, there are a larger number of asymptomatic cases in 
any epidemic, which nevertheless acquire immunity. It is estimated that for 
every case reported, 10-20 are not seen by health authorities (personnal 
communication).
Thirdly, under-reporting is a serious issue in Brazil. 
Duarte and Fran\c{c}a\cite{duarte_data_2006}, estimated 
the sensitivity of dengue reporting for hospitalized patients in 
Belo-Horizonte, Brazil to be of 63\%, 
meaning that aproximately 37\% of the suspected dengue cases go unreported.  
Lastly, demography and migrations affect the number of susceptible in ways which 
are not easy to determine.

The initial number of susceptibles -- $S_0$ -- is a a very powerful parameter 
with regards to epidemic risk.
Its value affects directly the local $R_0$, thus being intrumental in predicting 
epidemics, but it also affects the speed and size of the epidemic.

In this paper, we estimate the number of susceptibles, allowing for 
discontinuities right before every epidemic in the last 18 years. 
Through this method we hope to detect drastic changes in the mixture of 
susceptibles cause by the replacement of the predominant serotype, which 
can redefine susceptibility for the following season.

@@

\bibliography{lm1}
\section*{Figure Legends}
% This section is for figure legends only, do not include
% graphics in your manuscript file.
%
%\begin{figure}
%\caption{
%{\bf Bold the first sentence.}  Rest of figure caption.  
%}
%\label{Figure_label}
%\end{figure}
\begin{figure}
\begin{center}
\includegraphics[width=.9\textwidth]{./plots/concat_SI.png}
\end{center}
\caption{
{\bf Bold the first sentence.}  Rest of figure caption.  
}
\label{Fig:S0}
\end{figure}

@@

\begin{table}
\caption{\bf{Table title}}
 \begin{center}
            \begin{tabular}{c|c}
            \hline
            Year & Ratio \\
            \hline
            1997 & 204.26(67.45-706.68)\\
1998 & 207.45 (170.45-261.63)\\
2000 & 2300.63 (1741.30-3156.85)\\
2001 & 317.62 (226.47-526.45)\\
2005 & 100.28 (48.95-171.58)\\
2006 & 198.26 (114.59-317.88)\\
2007 & 567.78 (388.55-834.58)\\
2008 & 7.82 (2.97-34.43)\\
1999 & 26.70 (5.97-87.36)\\

        \hline
        \end{tabular}
        \end{center}
\end{table}
